---
title: "Halloween MiniProject"
author: "Diana Furlan"
date: "2024-10-28"
output: pdf_document
---

##1. Importing candy data



```{r}

url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy_file <- read.csv(url)
candy = read.csv(url, row.names=1)
head(candy)

```

##What is in the dataset?

> Q1. How many different candy types are in this dataset?

```{r}
candies <- nrow(candy)
candies
```

> Q2. How many fruity candy types are in the dataset?

```{r}
fruity <- sum(candy$fruity == 1)
fruity
```

##What is your favorate candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Almond Joy", ]$winpercent

```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```
There is a useful'skim()'function in the skimr package that can help give you a quick overview of a given dataset

```{r}

#install.packages("skimr")
library("skimr")
#or just extract a part from the library 
#skimr::skim(candy)
skim(candy)
```
we have to scale the winpercent

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset? sugarpercent, pricepercent and winpercent are on a non binary scale.

> Q7. What do you think a zero and one represent for the candy\$chocolate column? 1 = T for chocolate and 0 = F for chocolate

-------------------------------------
>Q8. Plot a histogram of winpercent values

```{r}
#can add color with 'col'- how to change the title?

hist(candy$winpercent, xlab = "Win Percent")
  abline(v = 50, col = "red")
```


```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth = 8) 
```

>Q9. Is the distribution of winpercent values symmetrical? slighly right-skewed


>Q10. Is the center of the distribution above or below 50%?
slightly above 50% 

```{r}
summary(candy$winpercent)
```


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?
chocolate seems higher

```{r}
summary(candy[as.logical(candy$chocolate),]$winpercent)
```
```{r}
#installed.packages("dplyr")
```
```{r}
library(dplyr)
```


```{r}
fruit.candy <- candy |>
   filter(fruity==1)

summary(fruit.candy$winpercent)

```

>Q12. Is this difference statistically significant?

```{r}
choc.candy <- candy |> filter(chocolate ==1)
fruit.candy <- candy |> filter(fruity == 1)
```

```{r}
t_test_result <- t.test(choc.candy$winpercent, fruit.candy$winpercent)

t_test_result
```
##Overall Candy Rankings


```{r}
play <- c("d","a","c")
sort(play)
order(play)

play[order(play) ]
```
>Q13. What are the five least liked candy types in this set?

```{r}
sort(c(5, 2, 10), decreasing = T)
```


```{r}
head( candy[order(candy$winpercent),], 5)

```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy%>%
  arrange(winpercent) %>% head(5)
```


>Q15. Make a first barplot of candy ranking based on winpercent values.

Bar plot 
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

Bar plot 
```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent),
      fill= chocolate) +
  geom_col()
```


More custom col skim  so we can see both chocolate and bar and fruity, etc. all from the same plot

```{r}
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols [as.logical(candy$bar)] <- "brown"
mycols [as.logical(candy$fruity)] <- "pink"
#use blue for my fav candy
```

```{r}
mycols [rownames(candy)=="Almond Joy"] <- "blue"
mycols
```

```{r}
#placeholder
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent),) +
  geom_col(fill =mycols)
```
>Q17. What is the worst ranked chocolate candy? Nik L Nip

>Q18. What is the best ranked fruity candy? Starbust

Plot of win vs Price
```{r}
mycols[as.logical(candy$fruity)]<- "red"
mycols
```

##Taking a look at pricepercent

add labels
```{r}
#install.packages("ggrepel")
```

#```{r}
library(ggrepel)
ggplot (candy) +
  aes(winpercent,pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_label(col=mycols) +
  geom_text_repel(max.overlaps = 8, col=mycols, ) 
#```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ggplot (candy)+
  aes(winpercent,pricepercent)+
  geom_point(col=mycols)

```

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head ( candy[ ord, c(11,12)], n=5 )
```



```{r}
#install.packages("corrplot")
```

```{r}
library(corrplot)
cij <- cor(candy)
corrplot (cij, diag=F)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
chocolate and gruit


>Q23. Similarly, what two variables are most positively correlated? chocolate and bar


##Principal Component Analysis

```{r}

pca <- prcomp(candy, scale= T)
summary(pca)
```
```{r}
plot(pca$x[,1], pca$x[,2], col=mycols, pch=16, xlab = "PC1", ylab = "PC2")
```
how do the original variables cols contribute to the new pca. PC1:
```{r}
loadings <- cbind(candy, pca$x[,1:3])
```


```{r}

p <- ggplot(loadings) +
  aes(x= PC1, y= PC2, 
      text=rownames(loadings), fill = PC1, size= winpercent/100) +
  geom_point(col=mycols)
  
  p
```

```{r}
#install.packages("ggrepel")
library(ggrepel)

q <- ggplot(loadings) +
  aes(x = PC1, y = PC2, text = rownames(loadings), fill = PC1, size = winpercent / 100) +
  geom_point(col=mycols) +
  geom_text_repel(aes(label = rownames(loadings)), size = 3.3, col = mycols, max.overlaps = Inf) + 
  theme(legend.position = "none") +
  labs(title = "Halloween Candy PCA Space", subtitle = "Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
    caption = "Data from 538")
q

```


```{r}
#install.packages("plotly")
#library(plotly)

#ggplotly(q)

```



```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?No,I think it should be reveresed